---
title: "Processing SMarTaR-ID reports"
author: "Deep Sea Conservation Research Unit - Nils Piechaud & Kerry Howell"
date: "2022"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: github

---

This document is to help you convert your annotated data downloaded from Biigle into a format that is suitable for multivariate analysis. It also enables you to link the annotation labels to taxonomic data in the World Register of Marine Species https://www.marinespecies.org/, using their aphia ID numbers.     

## Check that necessary packages have been installed

```{r, message = FALSE}

install.packages( 
setdiff( c("tidyverse","magrittr"),installed.packages()[,1] )
)

library(tidyverse)
library(magrittr)

```

## Set the working environment ## 

This will create the folder needed for the rest of this code to work properly

You can replace the path to the location of the folder you want to work from on your machine (C:/User/Name/Documents/...)
Or leave the code as it is, the first line will pick up the pathway to where your file is


```{r, message=FALSE , warning=FALSE}
# fetch the pathway to the dir where your script is
wd <- dirname(rstudioapi::getSourceEditorContext()$path)

paste0(wd,"/reports") -> reports_dir

paste0(wd,"/taxonomy") -> taxonomy_dir
 
```

Now that the folders are ready you can manually drop the Biigle CSV reports zips (as many as you want) in the 'reports' folder. 
Instructions to download the csv reports can be found here:
https://biigle.de/manual/tutorials/reports/reports-schema#annotation-csv-report

You need to select the csv type of 'Report variant' in Biigle. 
If you download reports from Biigle at the level of the project, you will need to first unzip the main folder, leaving you with multiple zipped volume folders to the add to the 'reports' folder.

Check that the sheet "Label_tree_aphiaIDs.csv" has successfully downloaded into the 'Taxonomy' folder. This sheet is used to add the aphiaIDs to the smartrID names. If it has successfully installed itself in the folder, all is well, you may continue running code. 

If it is not present in the Taxonomy folder, you will need to manually download from here: https://github.com/DeepSeaCRU/CRU-resources/blob/main/Label_tree_aphiaIDs.csv, and then add to the folder 'Taxonomy'. If the sheet "Label_tree_aphiaIDs.csv" cannot be found, the script will throw an error.

When that is done you can run the rest of the script.


```{r}


# add the taxonomy to the lables (so annotations can be grouped into higher levels bins)
paste0(taxonomy_dir,"/Label_tree_aphiaIDs.csv") %>% read_csv() %>% 
  select(- label_tree_id) -> aphiaIDs


# Open all biigle CSV reports in reports_dir

# list the csv tables -
list.files(reports_dir, pattern = "csv_image_annotation_report") -> files

# 2) make a metadata table --------------------------------------------------------------------------------

tibble(file = files) %>%
  mutate(table_name = str_remove(file, pattern = ".zip")) %>%
  # make a column of volume ID number
  mutate(volume = str_remove(table_name, pattern = "_csv_image_annotation_report")) -> annotationTables


# make a list to store all transformed tables tables
Dframes  <- as.list(1:nrow(annotationTables))

for (i in seq(Dframes)) {
  # select the table number i
  annotationTables %>% slice(i) -> meta.i
  # read the name of the table 
 meta.i %>% pull(file) %>% paste(reports_dir, ., sep = "/") %>%
   zip::zip_list() %>%
   pull(filename) %>%
   str_remove(".csv") %>%
   str_remove(pattern = paste0(meta.i$volume,"-")) -> meta.i$volume_name
  # import it
  meta.i %>% pull(file) %>% paste(reports_dir, ., sep = "/") %>%
    read_csv(col_types = cols()) -> D.i
  
  # if there are no annotations, skip to the next report
  if(D.i %>% nrow() < 1){
    print(paste0("No annotations in ",meta.i$volume, "-", meta.i$volume_name))
    tibble() ->  Dframes[[i]]
    next
  }
  # add the metadata
  bind_cols(D.i, meta.i) ->   D.imeta
    
  D.imeta ->  Dframes[[i]]
  
  rm(D.i, D.imeta)
  
}

# This is your table of everything
Dframes %>% bind_rows() -> allannotations



```

## separate the taxonomic levels 
```{r}

# table of metadata per OTU
allannotations %>%  distinct(label_name, label_hierarchy) -> OTU_meta


# label hierarchy without the final OTU name
# number of hierarchy levels
OTU_meta %<>% mutate(n_taxonomy_levels = label_hierarchy %>%
                       str_split(pattern = " > ") %>%
                       map(function(x)
                         return(length(x))) %>%
                       unlist)
# make the taxonomy column a table
taxonomies <- OTU_meta %>%
  pull(label_hierarchy) %>%
  str_split(pattern = " > ") %>%
  map(function(x)
    return(head(x, -1))) # remove last element as it is the OTU
# add names to thins list
names(taxonomies) <- OTU_meta %>% pull(label_name)
# maximum number of levels
taxonomies %>% map(length) %>% unlist() %>% max() -> max_levels

# re-arrange table 
otutable <- function(x) {
  c(x  , rep("OTU", max_levels - length(x))) %>%
    as_tibble() %>%  mutate(levels = paste0("level", 1:max_levels)) %>%
    pivot_wider(names_from = levels)
}

# add it to the mapped 
taxonomies %<>% map(otutable) %>% bind_rows()

OTU_meta %<>%
  select(-label_hierarchy) %>%
  bind_cols(taxonomies)


# add the info from the catalogue to the OTU table
OTU_meta %<>% left_join(aphiaIDs, by = c("label_name" = "name"))


#add the OTU info to the annotations
allannotations %>% left_join(OTU_meta, by = "label_name") -> allannotations_taxonomy


```


## Rearrange tables and export to CSV

```{r}


allannotations_taxonomy %>% distinct(filename, volume) %>% count(filename) %>% filter(n > 1) -> duplicated_image_check

if( nrow(duplicated_image_check)> 0){ 
  print("Some of your images are present in multiple volumes -- volume names will be included in the matrices")
  
}


# rearrange it and calculate the count of each OTU per image
allannotations_taxonomy %>%
  # keep the label and its hierarchy
  group_by(image_id, filename, volume,volume_name, label_name, label_hierarchy) %>%
  # count the
  summarise(count = n()) %>%
  ungroup() %>%
  mutate(image_id = as.character(image_id)) -> allannotations_count

# rearrange the table into useful formats -------------------------------------------------------------------
# contingency table
if(nrow(duplicated_image_check)> 0) {
  # contingency table
  spXsite <-    allannotations_count %>%
    pivot_wider(
      id_cols = c(filename, volume, volume_name),
      names_from = label_name,
      values_from = count,
      values_fill = list(count = 0)
    )
} else{
  # contingency table
  spXsite <-    allannotations_count %>%
    pivot_wider(
      id_cols = c(filename),
      names_from = label_name,
      values_from = count,
      values_fill = list(count = 0)
    )
}
# export it as a table into your Working directory
spXsite %>% write_csv(paste0(wd, "/", "spXsite.csv"))


# add the taxonomny as head over the species names
OTU_meta %>% select(label_name, aphia_id, contains("level")) %>% select(-n_taxonomy_levels) %>%
  column_to_rownames("label_name") %>%
  t %>%
  as.data.frame %>%
  rownames_to_column("key") %>% rename(filename = key)  %>% 

bind_rows( spXsite %>%  mutate_if(is.integer, as.character)) -> head_spXsite

head_spXsite %>% write_csv(paste0(wd, "/", "taxonomy_spXsite.csv"))


if( nrow(duplicated_image_check) > 0){
  # list of sample
  siteXsp <-   allannotations_count %>%
    pivot_wider(
      names_from  = c(filename, volume, volume_name),
      values_from = count,
      values_fill = list(count = 0)
    ) 
}else{
  # list of sample
  siteXsp <-   allannotations_count %>%
    pivot_wider(
      names_from  = filename,
      id_cols = label_name,
      values_from = count,
      values_fill = list(count = 0)
    ) 
} 
# export it
siteXsp %>% write_csv(paste0(wd, "/", "siteXsp.csv"))


# add the taxonomy info
taxonomy_siteXsp <- siteXsp %>%
  left_join(OTU_meta, ., by = "label_name")
# export it
taxonomy_siteXsp %>% write_csv(paste0(wd, "/", "taxonomy_siteXsp.csv"))

```








